Bayesian calibration of a hybrid nuclear collision model\\ using $\boldsymbol p$-Pb and Pb-Pb data from the LHC

### 2018-10-22 referee_respons.txt

Thank you for the constructive criticisms and comments. We have significantly
revised the manuscript to improve its clarity and address several of your concerns.

Response to the specific concerns of the referee:

"A parton model of the proton has direct consequences not only for
heavy-ion collisions, but also for p+p collisions. The authors should
present predictions for p+p scattering using the most probable
parameters (or using the posterior distribution of parameters w, v,
n_c) for the differential elastic p+p scattering cross section. The
prediction can be compared to parametrized data at 5TeV. As a proxy
for the p+p data the parametrization from Phys. Rev. D88 (2013) 094019
can be used."

We agree that it would be interesting to explore the consequences of our
proton structure parameters on p-p observables. However, doing so properly
would involve significant work beyond the scope of the present study. We
therefore believe that such an analysis would be more appropriate for an
independent follow-up study. With this in mind, we've made all software, methods,
and data open source to the broader community.

"The posterior for n_c (the number of constituents) clearly shows
that the prior range is too constrained, a large part of the posterior
extends in the region n_c>9. I do not want to ask the authors to redo
the calculation with different prior, but a comment should be added in
text."

Thank you for pointing this out. We've added the following comment to
explain the truncation of the constituent number prior:

One exception is the constituent number nc which we limit for practical considerations.
Recall that each constituent fluctuates independently, weighted by a gamma random variable.
Hence for constituent numbers $n_c > 1$, the fluctuations average out, and the resulting nucleon fluctuations vanish.
To counteract this effect, the constituent fluctuation variance must increase as nc increases.
Eventually, these required fluctuations become unreasonably large.
We find that for nc < 10, the energy density fluctuations are reasonable, and hence limit nc to this prior range."

"For the model predictions from the posterior experimental data (even preliminary) should be added, whenever available."

We've been careful to include all available data (public or otherwise), but it
is possible that we missed a dataset. In response to your suggestion, I've sent an email to the ALICE
collaboration to request their preliminary data, e.g. Pb-Pb mean pT, which could be
added to the posterior predictions where data is missing. Unfortunately, I have not heard back from them
yet, and I do not believe it is their policy to share preliminary data.
If you can direct me to a specific dataset, I would be happy to include it in
the maximum a posteriori observable plot.

Additional changes to the manuscript:

1. We've replaced the title with a more succinct version.

2. Several technical details of the initialization procedure have been
rewritten based on comments received from the community. The initial condition
parametrization is correspondingly recast as an initial energy density profile
(previously it was a parton density). We've also improved the
discussion of the initial free-streaming evolution and its matching to viscous
hydrodynamics.

3. We've simplified the nucleon substructure parameters, replacing the nucleon
width w [fm] and nucleon structure parameter X with a constituent sampling
radius r and constituent width v [fm]. The new parametrization is simpler and
less prone to misinterpretaiton. This transformation does not change the
conclusions of the study (nor should it).

4. We've corrected several inaccuracies that we made when discussing the
posterior nucleon width. In particular, the RMS radius differs from the Gaussian
width by a factor of sqrt(3).

5.

### 2018-10-22 paper/paper.tex
Begin crafting referee response.

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_\mathrm{NN}}}
\newcommand{\fmc}{\ensuremath{\text{fm}/c}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\sigmaf}{\sigma_\text{fluct}}
\newcommand{\taufs}{\tau_\text{fs}}
\newcommand{\dmin}{d_\text{min}}
\newcommand{\dTb}{\Delta T_\text{bulk}}
\newcommand{\Tb}{T_\text{bulk}}
\newcommand{\Tsw}{T_\text{switch}}
\newcommand{\T}{\tilde{T}}
\newcommand{\bv}{\mathbf b}
\newcommand{\xv}{\mathbf x}
\newcommand{\yv}{\mathbf y}
\newcommand{\zv}{\mathbf z}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

Consider the collision of two protons, labeled $A$ and $B$, with three-dimensional densities $\rho_A$ and $\rho_B$ in their local rest frames.
The proton-proton overlap function
\begin{equation}
  \label{eq:tpp}
  T_{pp}(b) = \int dx\, dy \int dz_1\, \rho_A(x, y, z_1) \int dz_2\, \rho_B(x - b, y, z_2),
\end{equation}
describes the eikonal overlap of the two proton wave packets at fixed impact parameter $b$, where coordinate $(x, y)$ lies in the transverse plane and $z$ is parallel to the beam axis.
Here we assume that each proton is comprised of smaller constituents---e.g.\ valence quarks, sea quarks, and small-x gluons---which may collide to produce secondary particles and contribute to the observed inelastic proton-proton cross section.

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom is uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's three-dimensional density $\rho_{A,B}$ as a sum of $n_c$ independent constituents,
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\xv) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\xv - \xv_i),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution,
\begin{equation}
  \rho_c(\xv) = \frac{1}{(2 \pi v^2)^{3/2}} \exp\left(- \frac{|\xv|^2}{2 v^2}\right).
\end{equation}
of variable width $v$.
The constituent positions $\xv_i$ are sampled independently (ignoring correlations) from a Gaussian radial distribution,
\begin{equation}
  \label{eq:radial_dist}
  \rho_\text{pos}(\xv_i; \xv_{A,B}) = \frac{1}{(2\pi r^2)^{3/2}} \exp\left(-\frac{|\xv_i - \xv_{A,B}|^2}{2 r^2}\right),
\end{equation}
where $\xv_{A,B}$ is the position of each proton, and $r$ is a parameter which varies the sampling radius of the constituent positions about each proton position.
As a matter of convenience, we sample the proton positions \emph{before} determining the constituent positions.
This leads to small discrepancies between the designated proton positions $\xv_{A,B}$ and the actual position of each proton's center-of-mass, owing to fluctuations in the constituent positions $\xv_i$.
The sampling radius $r$ should thus be interpreted with care.
It is \emph{not} the root-mean-square radius of the sampled nucleons.

@@

The energy density $e$ deposited at midrapidity $\eta_s=0$ and proper time $\tau_0 \ll 1~\fmc$ is then some function,
\begin{equation}
  e(\xv_\perp, \eta_s=0, \tau_0) = f(\T_A, \T_B),
\end{equation}
of the local participant thicknesses $\T_A$ and $\T_B$.
A natural first guess for this mapping is an arithmetic mean,
\begin{equation}
  e(\xv_\perp, \eta_s=0, \tau_0) = \text{const} \times \frac{\T_A + \T_B}{2},
\end{equation}
equal to a participant or ``wounded nucleon'' model for initial energy deposition up to meaningless factor of two in the overall normalization constant.
This wounded nucleon scaling was in fact one of the first such mappings used as a proxy for initial particle production in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was subsequently realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality \cite{Kharzeev:2000ph} and hence the wrong scaling for initial energy deposition at midrapidity as a function of participant thicknesses $\T_A$ and $\T_B$\footnote{Typically an entropy density $s$ is used, but the shortcomings of the wounded nucleon model are qualitatively the same when depositing energy.}.

@@

We note that previous applications of the \trento\ model \cite{Moreland:2014oya, Bernhard:2016tnd} used the generalized mean ansatz to parametrize the \emph{entropy} density $s(\xv_\perp, \eta_s=0, \tau_\text{hydro})$, motivated (in part) by the approximate linear relation at midrapidity between initial entropy density $dS/d\eta_s$ and final charged-particle density $d\nch/d\eta_s$ \cite{Shen:2015qta}.
These studies, however, assumed instant thermalization, an approximation which sets $u^\mu$, $\Pi$, and $\pi^{\mu\nu}$ identically to zero at the hydrodynamic starting time $\tau_\text{hydro}$.
The present study, meanwhile, uses a pre-equilibrium free streaming stage to estimate $u^\mu$, $\Pi$, and $\pi^{\mu\nu}$ as a function of a variable free-streaming time $\taufs$, at which point the system is matched to hydrodynamics.
Strictly speaking, the system's entropy is ill defined prior to thermalization while the system free streams.
We therefore reinterpret the quantity produced by the \trento\ model as an energy density $e(\xv_\perp, \eta_s=0, \tau_0)$ and free-stream this quantity to the hydrodynamic starting time $\tau_\text{hydro} \equiv \taufs$.
This assumes, of course, that the model's parametrization for entropy density is also suitable as a parametrization for the energy density.
To this point, we note that the arguments for the generalized mean ansatz are quite general \cite{Moreland:2015oya}, and that the energy density $dE/d\eta_s$ and entropy density $dS/d\eta_s$ share many similar qualities.
For example, both are approximately conserved at midrapidity, and both scale approximately linearly with the final charged-particle density $d\nch/d\eta_s$.
We thus expect the parametrization to be suitable for either quantity.

@@

We parametrize the temperature dependence of the QGP viscosities in order to marginalize over their uncertainty when calibrating to data.
For the specific shear viscosity $\eta/s$, we use a modified linear ansatz
\begin{equation}
  \label{eq:shear_viscosity}
  (\eta/s)(T) = (\eta/s)_\mathrm{min} + (\eta/s)_\mathrm{slope}\cdot(T - T_c)\cdot(T/T_c)^{(\eta/s)_\mathrm{crv}},
\end{equation}
where $\eta/s$ min, slope, and curvature are tunable parameters, and $T_c=0.154$~GeV is the pseudocritical transition temperature of the HotQCD EoS.
While for the specific bulk viscosity $\zeta/s$, we use an unnormalized Cauchy distribution
\begin{equation}
  \label{eq:bulk_viscosity}
  (\zeta/s)(T) = \frac{(\zeta/s)_\text{max}}{1 + \left(\dfrac{T - T_\text{bulk}}{\Delta T_\text{bulk}}\right)^2},
\end{equation}
described by a tunable maximum $(\zeta/s)_\text{max}$, width $\dTb$, and location parameter $\Tb$.
Figure~\ref{fig:viscosity_dof} shows several of the possible curves parametrized by Eqs.~\eqref{eq:shear_viscosity} and \eqref{eq:bulk_viscosity}, although many more are possible.

@@

\begin{table}[t]
  \caption{Input parameter ranges for the physics model.}
  \begin{ruledtabular}
  \begin{tabular}{lll}
    Parameter         & Description                        & Range             \\
    \paddedhline
    Norm              & Normalization factor                 & 9--28           \\
    $p$               & Entropy deposition parameter         & $-1$ to $+1$    \\
    $\sigmaf$         & Nucleon fluctuation std.\ dev.\      & 0--2            \\
    $r$               & Constituent sampling radius          & 0--1.2 fm       \\
    $n_c$             & Number of nucleon constituents       & 1--9            \\
    $v$               & Constituent width                    & 0.2--1.2 fm     \\
    $\dmin$           & Minimum inter-nucleon distance       & 0--1.7 fm       \\
    $\taufs$          & Free-streaming time                  & 0.1--1.5 \fmc   \\
    $\eta/s$ min      & Minimum value of $\eta/s$ (at $T_c$) & 0--0.2          \\
    $\eta/s$ slope    & Slope of $\eta/s$ above $T_c$        & 0--8 GeV$^{-1}$ \\
    $\eta/s$ crv      & Curvature of $\eta/s$ above $T_c$    & $-1$ to $+1$    \\
    $\zeta/s$ max     & Maximum value of $\zeta/s$           & 0--0.1          \\
    $\dTb$            & Width of $\zeta/s$ peak              & 0--0.1 GeV      \\
    $\Tb$             & Temperature of $\zeta/s$ maximum     & 150--200 MeV    \\
    $\Tsw$            & Switching/particlization temp.       & 135--165 MeV    \\
  \end{tabular}
  \end{ruledtabular}
  \label{tab:design}
\end{table}

@@

However, for some observables such as $p$-Pb mean $p_T$ \cite{Abelev:2013bla} and flow cumulants $\vnk{n}{k}$ \cite{Chatrchyan:2013nka}, the experiments use a special high-multiplicity trigger to select rare, ultra-central events according to the number of charged-particles produced $\nch$ or detector tracks offline $\ntrk$.
These high-multiplicity bins are too selective for our modest minimum bias event sample, and so a different procedure is required.
We exploit, for this purpose, the approximate monotonic relation between each event's initial transverse energy density,
\begin{equation}
  \frac{dE_T}{d\eta_s} \bigg\vert_{\eta_s=0} = \tau_0 \int d^2x_\perp e(\xv_\perp, \eta_s=0, \tau_0).
\end{equation}
and its final charged-particle density $(d\nch/d\eta) \vert_{\eta=0}$ at midrapidity.

Consider, for example, a single multiplicity bin $[\nch^\text{low}, \nch^\text{high}]$ which selects events from a minimum bias event sample with $p_T^\text{min} < p_T < p_T^\text{max}$ and $|\eta| < \Delta \eta$.
Let $\langle\nch\rangle$ denote the average charged-particle multiplicity of these events.
We first rescale the experimental multiplicity bin edges,
\begin{equation}
  [\nch^\text{low}, \nch^\text{high}] \rightarrow \left[\frac{\nch^\text{low}}{\langle \nch \rangle},\, \frac{\nch^\text{high}}{\langle \nch \rangle}\right ],
\end{equation}
in order to reexpress each bin edge as a unitless variable.
These bin edges are then associated with a pair of transverse energy bin edges,
\begin{equation}
  \label{eq:mult_trigger}
  \left [\frac{E_T^\text{min}}{\langle E_T \rangle}, \frac{E_T^\text{max}}{\langle E_T \rangle} \right] \leftrightarrow \left [\frac{\nch^\text{low}}{\langle \nch \rangle}, \frac{\nch^\text{high}}{\langle \nch \rangle} \right ],
\end{equation}
where $E_T \propto (dE_T/d\eta_s) \vert_{\eta_s=0}$ is the midrapidity transverse energy of a single event, and $\langle E_T \rangle$ is the corresponding average over the minimum bias event sample.

@@

\begin{table}[t]
  \caption{
    \label{tab:post_param}
    Posterior parameter estimates corresponding to Fig.~\ref{fig:posterior}.
    The reported values are for the distribution median and 90\% highest posterior density credible interval.
  }
  \begin{ruledtabular}
    \newlength{\cellwidth}
    \settowidth{\cellwidth}{$-0.00$}
    \newcommand{\est}[3]{\parbox{\cellwidth}{\hfill$#1$}$_{-#2}^{+#3}$}
    \begin{tabular}{llll}
      \toprule
      \multicolumn{2}{c}{Initial condition / Pre-eq}     & \multicolumn{2}{c}{QGP medium}              \\
      \cmidrule(r){1-2}                                    \cmidrule(l){3-4}
      \addlinespace[.4ex]
      Norm       & \est{20.0}{2.4}{2.4}          & $\eta/s$ min     & \est{0.10}{0.07}{0.07}            \\[1.1ex]
      $p$        & \est{0.013}{0.155}{0.144}     & $\eta/s$ slope   & \est{1.11}{1.11}{1.42} GeV$^{-1}$ \\[1.1ex]
      $\sigmaf$  & \est{0.93}{0.27}{0.26}        & $\eta/s$ crv     & \est{-0.14}{0.86}{0.81}           \\[1.1ex]
      $r$        & \est{0.88}{0.23}{0.25} fm     & $\zeta/s$ max    & \est{0.026}{0.026}{0.033}         \\[1.1ex]
      $n_c$      & \est{6.3}{3.1}{2.7}           & $\dTb$           & \est{0.033}{0.033}{0.048} GeV     \\[1.1ex]
      $v$       & \est{0.51}{0.18}{0.26} fm      & $\Tb$            & \est{0.176}{0.020}{0.024} GeV     \\[1.1ex]
      $\dmin$    & \est{1.13}{0.51}{0.55} fm     & $\Tsw$           & \est{0.149}{0.013}{0.014} GeV     \\[1.1ex]
      $\taufs$   & \est{0.37}{0.27}{0.33} \fmc   & \\
      \addlinespace[.4ex]
      \bottomrule
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

The \trento\ normalization factor ${\text{Norm} = 20.2^{+2.2}_{-2.2}}$ and energy deposition parameter $p=0.013^{+0.144}_{-0.155}$ are well constrained by the present analysis.
Moreover, Figs.~\ref{fig:obs_pbpb} and \ref{fig:obs_ppb} show that the model predictions using these values nicely describe both the $p$-Pb and Pb-Pb calibration observables.
While it would not be surprising to fit one or two of these observables using a narrow range of parameter values, the quality of the combined fit (more on this later) and the number of observables described is highly non-trivial.
For example, consider the ratio of the $p$-Pb charged-particle yield to the Pb-Pb charged-particle yield.
As the energy deposition parameter $p$ trends toward positive (negative) infinity, particle production scales like the maximum (minimum) of local nuclear overlap density.
This has a much stronger effect on the highly asymmetric $p$-Pb system than it does on the Pb-Pb system; hence the parameter $p$ strongly affects the ratio of the two average yields.

@@

Continuing down the diagonal in Fig.~\ref{fig:posterior}, we see that the constituent sampling radius $r=0.88_{-0.23}^{+0.25}$~fm, and the constituent width $v=0.51_{-0.18}^{+0.26}$~fm.
Their joint posterior distribution is shown in Fig.~\ref{fig:posterior_proton_shape}, illustrating the constraining power of the analysis.
Recall that the parameter $r$ controls the dispersion of the constituent positions, while $v$ varies each constituent's Gaussian width.
We emphasize that $r$ is model parameter, not a physical one.
It describes the radius of the constituent position sampling distribution, \emph{not} the average proton radius, a quantity that's measured in the proton's center-of-mass frame.
For example, when $n_c=1$, our model describes each nucleon as a single Gaussian of width $v$, yet the parameter $r$ could nevertheless assume any value within its prior.
Despite this idiosyncrasy, one can easily calculate a physical nucleon width \emph{ex post facto}, given specific values for $r$, $v$, and $n_c$.

@@

The posterior free-streaming time $\taufs=0.37_{-0.27}^{+0.33}\ \fmc$ obtained in this work is significantly smaller than the previous estimate $\taufs=1.16_{-0.25}^{+0.29}\ \fmc$ quoted in Ref.~\cite{Bernhard:2018hnz}.
We point out that the present study is missing several important observables which could affect the estimated free-streaming time, e.g.\ the Pb-Pb mean $p_T$ and mean $p_T$ fluctuations at $\sqrts=5.02$~TeV.
Nevertheless, it appears that the inclusion of nucleon substructure significantly reduces the maximum allowed free-streaming time, although more work is needed to establish if this is indeed the case.

>>>>

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_\mathrm{NN}}}
\newcommand{\fmc}{\ensuremath{\text{fm}/c}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\sigmaf}{\sigma_\text{fluct}}
\newcommand{\taufs}{\tau_\text{fs}}
\newcommand{\dmin}{d_\text{min}}
\newcommand{\Tsw}{T_\text{switch}}
\newcommand{\T}{\tilde{T}}

\newcommand{\smin}{{(\eta/s)_\mathrm{min}}}
\newcommand{\sslope}{{a_\mathrm{shear}}}
\newcommand{\scrv}{{c_\mathrm{shear}}}
\newcommand{\bmax}{{(\zeta/s)_\mathrm{max}}}
\newcommand{\bloc}{{T_\text{bulk}}}
\newcommand{\bwidth}{{\Delta T_\text{bulk}}}

\newcommand{\bv}{\mathbf b}
\newcommand{\xv}{\mathbf x}
\newcommand{\yv}{\mathbf y}
\newcommand{\zv}{\mathbf z}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

Consider the collision of two protons, labeled $A$ and $B$, with three-dimensional densities $\rho_A$ and $\rho_B$ in their local rest frames.
The proton-proton overlap function
\begin{equation}
  \label{eq:tpp}
  T_{pp}(b) = \int dx\, dy \int dz_1\, \rho_A(x, y, z_1) \int dz_2\, \rho_B(x - b, y, z_2),
\end{equation}
describes the eikonal overlap of the two proton wave packets at fixed impact parameter $b$, where coordinate $(x, y)$ lies in the transverse plane and $z_{1,2}$ is parallel to the beam axis.
Here we assume that each proton is comprised of smaller constituents---e.g.\ valence quarks, sea quarks, and small-x gluons---which may collide to produce secondary particles and contribute to the observed inelastic proton-proton cross section.

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom is uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's three-dimensional density $\rho_{A,B}$ as a sum of $n_c$ independent constituents,
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\xv) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\xv - \xv_i),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution,
\begin{equation}
  \rho_c(\xv) = \frac{1}{(2 \pi v^2)^{3/2}} \exp\left(- \frac{|\xv|^2}{2 v^2}\right).
\end{equation}
of variable width $v$.
The constituent positions $\xv_i$ are sampled independently (ignoring correlations) from a Gaussian radial distribution,
\begin{equation}
  \label{eq:radial_dist}
  \rho_\text{pos}(\xv_i; \xv_{A,B}) = \frac{1}{(2\pi r^2)^{3/2}} \exp\left(-\frac{|\xv_i - \xv_{A,B}|^2}{2 r^2}\right),
\end{equation}
where $\xv_{A,B}$ is the position of each proton, and $r$ is a parameter which varies the sampling radius of the constituent positions about each proton position.
As a matter of convenience, we sample the proton positions \emph{before} determining the constituent positions.
This leads to small discrepancies between the designated proton positions $\xv_{A,B}$ and the actual position of each proton's center-of-mass, owing to fluctuations in the constituent positions $\xv_i$.
The sampling radius $r$ should thus be interpreted with care.
It is \emph{not} the Gaussian width of the sampled nucleons in their center-of-mass frame.

@@

The energy density $e$ deposited at midrapidity $\eta_s=0$ and proper time $\tau_0 \ll 1~\fmc$ is then some function,
\begin{equation}
  e(\xv_\perp, \eta_s=0, \tau_0) = f(\T_A, \T_B),
\end{equation}
of the local participant thicknesses $\T_A$ and $\T_B$.
A natural first guess for this mapping is an arithmetic mean,
\begin{equation}
  e(\xv_\perp, \eta_s=0, \tau_0) = \text{const} \times \frac{\T_A + \T_B}{2},
\end{equation}
equal to a participant or ``wounded nucleon'' model for initial energy deposition up to meaningless factor of two in the overall normalization constant.
This wounded nucleon scaling was in fact one of the first such mappings used as a proxy for initial particle production in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was subsequently realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality \cite{Kharzeev:2000ph} and hence the wrong scaling for initial midrapidity energy\footnote{Typically an entropy density $s$ is used, but the shortcomings of the wounded nucleon model are qualitatively the same the energy density $e$ is used instead.} deposition as a function of participant density.

@@

We note that previous applications of the \trento\ model \cite{Moreland:2014oya, Bernhard:2016tnd} used the generalized mean ansatz to parametrize the \emph{entropy} density $s(\xv_\perp, \eta_s=0, \tau_\text{hydro})$, motivated (in part) by the approximate linear relation at midrapidity between initial entropy density $dS/d\eta_s$ and final charged-particle density $d\nch/d\eta_s$ \cite{Shen:2015qta}.
These studies, however, assumed instant thermalization, an approximation which sets $u^\mu$, $\Pi$, and $\pi^{\mu\nu}$ identically to zero at the hydrodynamic starting time $\tau_\text{hydro}$.
The present study, meanwhile, uses a pre-equilibrium free streaming stage to estimate $u^\mu$, $\Pi$, and $\pi^{\mu\nu}$ as a function of a variable free-streaming time $\taufs$, at which point the system is matched to hydrodynamics.
Strictly speaking, the system's entropy is ill defined prior to thermalization while the system free streams.
We therefore reinterpret the quantity produced by the \trento\ model as an energy density $e(\xv_\perp, \eta_s=0, \tau_0)$ and free-stream this quantity to the hydrodynamic starting time $\tau_\text{hydro} \equiv \taufs$.
This assumes, of course, that the model's parametrization for entropy density is also suitable as a parametrization for the energy density.
To this point, we note that the arguments for the generalized mean ansatz are quite general \cite{Moreland:2014oya}, and that the energy density $dE/d\eta_s$ and entropy density $dS/d\eta_s$ share many similar qualities.
For example, both are approximately conserved at midrapidity, and both scale approximately linearly with the final charged-particle density $d\nch/d\eta_s$.
We thus expect the parametrization to be suitable for either quantity.

@@

We parametrize the temperature dependence of the QGP viscosities in order to marginalize over their uncertainty when calibrating to data.
For the specific shear viscosity $\eta/s$, we use a modified linear ansatz
\begin{equation}
  \label{eq:shear_viscosity}
  (\eta/s)(T) = \smin + \sslope \cdot(T - T_c)\cdot \left ( \frac{T}{T_c} \right)^\scrv,
\end{equation}
where $\smin$, $\sslope$, and $\scrv$ are tunable parameters, and $T_c=0.154$~GeV is the pseudocritical transition temperature of the HotQCD EoS.
While for the specific bulk viscosity $\zeta/s$, we use an unnormalized Cauchy distribution
\begin{equation}
  \label{eq:bulk_viscosity}
  (\zeta/s)(T) = \frac{\bmax}{1 + \left(\dfrac{T - \bloc}{\bwidth}\right)^2},
\end{equation}
described by a tunable maximum $\bmax$, temperature width $\bwidth$, and temperature location parameter $\bloc$.
Figure~\ref{fig:viscosity_dof} shows several of the possible curves parametrized by Eqs.~\eqref{eq:shear_viscosity} and \eqref{eq:bulk_viscosity}, although many more are possible.

@@

\begin{table}[t]
  \caption{Input parameter ranges for the physics model.}
  \begin{ruledtabular}
  \begin{tabular}{lll}
    Parameter         & Description                        & Range             \\
    \paddedhline
    Norm              & Normalization factor                 & 9--28 GeV       \\
    $p$               & Energy deposition parameter         & $-1$ to $+1$     \\
    $\sigmaf$         & Nucleon fluctuation std.\ dev.\      & 0--2            \\
    $r$               & Constituent sampling radius          & 0--1.2 fm       \\
    $n_c$             & Number of nucleon constituents       & 1--9            \\
    $v$               & Constituent width                    & 0.2--1.2 fm     \\
    $\dmin$           & Minimum inter-nucleon distance       & 0--1.7 fm       \\
    $\taufs$          & Free-streaming time                  & 0.1--1.5 \fmc   \\
    $\smin$           & Minimum value of $\eta/s$ (at $T_c$) & 0--0.2          \\
    $\sslope$         & Slope of $\eta/s$ above $T_c$        & 0--8 GeV$^{-1}$ \\
    $\scrv$           & Curvature of $\eta/s$ above $T_c$    & $-1$ to $+1$    \\
    $\bmax$           & Maximum value of $\zeta/s$           & 0--0.1          \\
    $\bwidth$         & Width of $\zeta/s$ peak              & 0--0.1 GeV      \\
    $\bloc$           & Temperature of $\zeta/s$ maximum     & 150--200 MeV    \\
    $\Tsw$            & Switching/particlization temp.       & 135--165 MeV    \\
  \end{tabular}
  \end{ruledtabular}
  \label{tab:design}
\end{table}

@@

However, for some observables such as $p$-Pb mean $p_T$ \cite{Abelev:2013bla} and flow cumulants $\vnk{n}{k}$ \cite{Chatrchyan:2013nka}, the experiments use a special high-multiplicity trigger to select rare, ultra-central events according to the number of charged-particles produced $\nch$ or detector tracks offline $\ntrk$.
These high-multiplicity bins are too selective for our modest minimum bias event sample, and so a different procedure is required.
We exploit, for this purpose, the approximate monotonic relation between each event's initial transverse energy density,
\begin{equation}
  \frac{dE_T}{d\eta_s} \bigg\vert_{\eta_s=0} = \tau_0 \int d^2x_\perp e(\xv_\perp, \eta_s=0, \tau_0),
\end{equation}
and its final charged-particle density $(d\nch/d\eta) \vert_{\eta=0}$ at midrapidity.

Consider, for example, a single multiplicity bin $[\nch^\text{low}, \nch^\text{high}]$ which selects events from a minimum bias event sample with $p_T^\text{min} < p_T < p_T^\text{max}$ and $|\eta| < \eta^\text{max}$.
Let $\langle\nch\rangle$ denote the average charged-particle multiplicity of these events.
We first rescale the experimental multiplicity bin edges,
\begin{equation}
  [\nch^\text{low}, \nch^\text{high}] \rightarrow \left[\frac{\nch^\text{low}}{\langle \nch \rangle},\, \frac{\nch^\text{high}}{\langle \nch \rangle}\right ],
\end{equation}
in order to reexpress each bin edge as a unitless variable.
These bin edges are then associated with a pair of transverse energy bin edges,
\begin{equation}
  \label{eq:mult_trigger}
  \left [\frac{E_T^\text{min}}{\langle E_T \rangle}, \frac{E_T^\text{max}}{\langle E_T \rangle} \right] \leftrightarrow \left [\frac{\nch^\text{low}}{\langle \nch \rangle}, \frac{\nch^\text{high}}{\langle \nch \rangle} \right ],
\end{equation}
where $E_T \propto (dE_T/d\eta_s) \vert_{\eta_s=0}$ is the midrapidity transverse energy of a single event in the desired kinematic range, and $\langle E_T \rangle$ is the corresponding average transverse energy over the full minimum bias event sample.

@@

\begin{table}[t]
  \caption{
    \label{tab:post_param}
    Posterior parameter estimates corresponding to Fig.~\ref{fig:posterior}.
    The reported values are for the distribution median and 90\% highest posterior density credible interval.
  }
  \begin{ruledtabular}
    \newlength{\cellwidth}
    \settowidth{\cellwidth}{$-0.00$}
    \newcommand{\est}[3]{\parbox{\cellwidth}{\hfill$#1$}$_{-#2}^{+#3}$}
    \begin{tabular}{llll}
      \toprule
      \multicolumn{2}{c}{Initial condition / Pre-eq}     & \multicolumn{2}{c}{QGP medium}              \\
      \cmidrule(r){1-2}                                    \cmidrule(l){3-4}
      \addlinespace[.4ex]
      Norm       & \est{20.0}{2.5}{2.6} GeV      & $\smin$     & \est{0.08}{0.07}{0.07}            \\[1.1ex]
      $p$        & \est{0.002}{0.180}{0.157}     & $\sslope$   & \est{1.24}{1.24}{1.46} GeV$^{-1}$ \\[1.1ex]
      $\sigmaf$  & \est{0.91}{0.33}{0.32}        & $\scrv$     & \est{-0.09}{0.91}{0.80}           \\[1.1ex]
      $r$        & \est{0.88}{0.23}{0.26} fm     & $\bmax$     & \est{0.026}{0.026}{0.032}         \\[1.1ex]
      $n_c$      & \est{6.0}{3.4}{3.0}           & $\bwidth$   & \est{0.035}{0.035}{0.043} GeV     \\[1.1ex]
      $v$        & \est{0.52}{0.20}{0.28} fm     & $\bloc$     & \est{0.174}{0.024}{0.020} GeV     \\[1.1ex]
      $\dmin$    & \est{1.12}{0.49}{0.58} fm     & $\Tsw$      & \est{0.149}{0.014}{0.013} GeV     \\[1.1ex]
      $\taufs$   & \est{0.47}{0.37}{0.55} \fmc   & \\
      \addlinespace[.4ex]
      \bottomrule
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

The \trento\ normalization factor ${\text{Norm} = 20.0^{+2.6}_{-2.5}}$ and energy deposition parameter $p=0.002^{+0.157}_{-0.180}$ are well constrained by the present analysis.
Moreover, Figs.~\ref{fig:obs_pbpb} and \ref{fig:obs_ppb} show that the model predictions using these values nicely describe both the $p$-Pb and Pb-Pb calibration observables.
While it would not be surprising to fit one or two of these observables using a narrow range of parameter values, the quality of the combined fit (more on this later) and the number of observables described is highly non-trivial.
For example, consider the ratio of the $p$-Pb charged-particle yield to the Pb-Pb charged-particle yield.
As the energy deposition parameter $p$ trends toward positive (negative) infinity, particle production scales like the maximum (minimum) of the two nuclear thickness functions.
This has a much stronger effect on the highly asymmetric $p$-Pb system than it does on the Pb-Pb system; hence the parameter $p$ strongly affects the ratio of the two average yields.

@@

Continuing down the diagonal in Fig.~\ref{fig:posterior}, we see that the constituent sampling radius $r=0.88_{-0.23}^{+0.26}$~fm, and the constituent width $v=0.52_{-0.20}^{+0.28}$~fm.
Figure~\ref{fig:posterior_proton_shape} shows the joint posterior distribution of both parameters, illustrating the constraining power of the Bayesian analysis.
While the sampling radius $r$ varies the size of the nucleons, we caution that its specific meaning should be interpreted with care; it specifies a computational \emph{sampling} radius, not a physical nucleon width.
Consider, for instance, a single nucleon with $n_c = 2$ constituents.
If the two constituents positions land on the same side of the nucleon, the effective nucleon size will be smaller than the Gaussian sampling radius $r$.
Despite this idiosyncrasy, one can easily define a physical nucleon width in the nucleon center-of-mass frame \emph{ex post facto}, given specific values for the sampling radius $r$, constituent width $v$, and constituent number $n_c$.

@@

The posterior free-streaming time $\taufs=0.47_{-0.37}^{+0.55}~\fmc$ obtained in this work is significantly smaller than the previous estimate $\taufs=1.16_{-0.25}^{+0.29}\ \fmc$ quoted in Ref.~\cite{Bernhard:2018hnz}.
We point out that the present study is missing several important observables which could affect the estimated free-streaming time, e.g.\ the Pb-Pb mean $p_T$ and mean $p_T$ fluctuations at $\sqrts=5.02$~TeV.
Nevertheless, it appears that the inclusion of nucleon substructure significantly reduces the maximum allowed free-streaming time, although more work is needed to establish if this is indeed the case.
